Article Title

### 2014-06-16 plos.tex
rev 53

We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best represented as the combination of a sparse matrix of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

Improved estimation of the correlation matrix is beneficial in itself. For example, improved estimates can be used to optimize  decoding of the population activity \cite{Friedman:1989}. But reduced estimation error is not the only benefit of regularization.  The amount of improvement of estimation is greatest when the structure imposed by regularization limits the solution space, but still represents the dominant dependencies in the data. Therefore, by finding the most efficient among many regularized estimators, we learn something about the system itself: the structure of the most efficient estimator is a parsimonious representation of the regularities in the data.

The necessity and benefits of regularization increase with the size of the recorded population. With the advent of  big neural data \cite{Alivisatos:2013}, the search for optimal regularization schemes will become increasingly relevant in any model of population activity. Since optimal regularization schemes are specific to systems under investigation, the inference of functional connectivity in large scale neural data will entail the search for optimal regularization schemes that may involve combinations of heuristic rules and numerical techniques specially designed for each neural circuit.

@@

We then performed a cross-validated evaluation to establish which of the four regularized estimators was most efficient for the population activity of dense groups of neurons in mouse primary visual cortex recorded with high-speed 3D random-access two-photon imaging of calcium signals. In our data, the sample correlation coefficients were largely positive and low.  We found that the best estimate of the correlation matrix was $C_{\sf sparse+latent}$.  This estimator revealed a sparse network of partial correlations (\sq{interactions}), between the observed neurons; it also inferred several latent units exerting linear effects on the observed neurons. We analyzed these networks of partial correlations and found the following: Whereas significant noise correlations were predominantly positive, the inferred interactions had a large fraction of negative values possibly reflecting inhibitory circuitry.  Moreover, we found that these interactions exhibited a stronger relationship to the physical distances and to the differences in preferred orientations than noise correlations. In contrast, the inferred negative interactions were less selective.

@@

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is an $p\times 1$ vector of firing rates in time bin $t$, and an independent unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix},
\begin{equation}\label{eq:sample}
	C_{\sf sample} = \left[\mbox{cov}\left(x(t)\right)\right]_{t\in T} =\frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
\end{equation}
where $n$ is the size of $T$, yields an unbiased estimate of the covariance matrix so that $\E{C_{\sf sample}}=\Sigma$.

@@

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominently in probablistic \emph{graphical modelling} wherein the joint distribution is explained by sets of two-way interactions \cite{Whittaker:1990}. For the multivariate Gaussian distribution, zero partial correlations indicate conditional independence of the pair, implying a lack of direct interaction between them \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful as indicators of conditional independence (lack of functinonal connectivity) between pairs of neurons.

Pairwise partial correlations are closely linked to the elements of the \emph{precision matrix}, \ie the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1990}. Zero elements in the precision matrix signify zero partial correlation between the two variables. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

As the size of recorded neuronal populations increases, without substantial increases in recording durations, the \emph{condition number} of the sample covariance matrix also increases, making the partial correlation estimates \emph{ill-conditioned}: small errors in covariance estimates translate into much larger errors in partial correlations \cite{Ledoit:2004,Schafer:2005}. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization.}

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure01.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
    {\bf Row 1.} Graphical representations of the target estimates of the four respective regularized covariance matrix estimators.  Recorded neurons are represented by green spheres and latent units by light-shaded spheres.  Edges represent non-zero partial correlations, \ie \sq{interactions}.
        {\bf Row 1, A}.  For estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear dependencies. 
        {\bf  Row 1, B.} For estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), representing systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. For estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from a sparse set of  linear associations between the observed units. 
        {\bf  Row 1, D}.  For estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to sparse associations between the recorded cells \emph{and} with several latent units. \\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with two latent units, {\bf C.}  a correlation matrix with 85\%  off-diagonal zeros in its inverse, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-1 (\ie one latent unit) matrix and a sparse matrix with 89\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=500$ drawn from simulated random processes with respective correlation matrices shown in Row 2.  The structure of the sample correlation matrix is difficult to discern.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:excess-loss}) for the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with structure that matches the true model converged to zero faster than the other estimators. Some exceptions occur when the sample is too small to reveal the structure in the data. In these cases estimators with simpler structures can be more efficient than estimators with true structure.
\\
{\bf Row 6:} Cross-validation losses for the five estimators relative to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:1}
\end{FPfigure}

The target estimate of estimator $C_{\sf diag}$ is the diagonal matrix $D$ containing estimates of neurons' variances. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf sample}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}\label{eq:c-diag}
C_{\sf diag} = (1-\lambda) C_{\sf sample} + \lambda D
\end{equation}
The structure imposed by $C_{\sf diag}$ favors (performs better  on) populations   with no linear associations between the neurons (\figref{1}{Row 1, A}).  If sample correlations arise from spurious effects, $C_{\sf diag}$ is expected to be more efficient than other estimators.

@@

Estimator $C_{\sf sparse+latent}$ is obtained by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component: 
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a low-rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (\figref{1}{Row 1, D}) \cite{Chandrasekaran:2010,Ma:2013}.

@@

\paragraph{Simulation}
We next demonstrate how the most efficient among different regularized estimators can reveal the structure of correlations.
We constructed four $50\times 50$ covariance matrices, each with structure that matched one of the four regularized estimators (\figref{1}{\,Row 2, A--D}, see Methods for details about how these matrices were generated).  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=500$), the sample correlation matrices were contaminated by sampling noise and their underlying structures were difficult to discern (\figref{1}{\,Row 3}).

@@

This loss function is related to the multivariate normal log-likelihood function $L(\Sigma|C_{\sf sample})$,
\begin{equation}
     \loss{X,Y} \equiv -\ln(2\pi) - \frac 2 p L(X|Y)
\end{equation}
Normalization by $p$ makes the values of the loss function comparable across different population sizes.

@@

The choice of the normal loss function is motivated by mathematical convenience. We expect that the main conclusions of our study will not change qualitatively with other well behaved loss functions, such as Stein's entropy loss or quadratic loss \cite{James:1961, Fan:2008, Ledoit:2004, Schafer:2005}.

The \emph{excess loss} function is produced by shifting the loss function to zero at its minimum:
\begin{equation}\label{eq:excess-loss}
    \eloss{C,\Sigma} = \loss{C,\Sigma}-\loss{\Sigma,\Sigma}
\end{equation}

We drew 30 independent samples of sizes $n=250$, 500, 1000, 2000, and 4000 from each model and computed the excess loss for each of the five estimators.  The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  All the regularized estimators produced better estimates (lower excess loss) than the sample covariance matrix.  However, estimators whose structure matched the true model typically outperformed the other estimators (\figref{1}{\,row 5}).  There were two exceptions: First, for some smaller samples, the data were insufficient to reveal the true correlation structure and the estimators with simpler structures, $C_{\sf diag}$ and $C_{\sf factor}$, often outperformed the estimators with the matching structure (\figref{1}{\,Row 5, C and D}).  Second, $C_{\sf sparse+latent}$ often performed as well as $C_{\sf sparse}$ on data from models without latent units because it reliably inferred the correct number of latent units: zero.

@@

With empirical data, the ground truth, $\Sigma$, is not accessible and excess loss cannot be computed directly. Instead, the loss function must be estimated from the sample through \emph{validation}.  In a validation procedure, an independent \emph{testing sample} is set aside to compute an additional sample covariance estimate $C_{\sf sample}^\prime$ to validate the estimate $C$ computed from the  \emph{training sample}.   The estimate $C_{\sf sample}^\prime$ is used to compute the \emph{validation loss} $\loss{C,C_{\sf sample}^\prime}$, which approximates the loss  $\loss{C,\Sigma}$.

The normal loss (Eq.~\ref{eq:loss}) is particularly convenient because it is additive in its second argument, \ie
\begin{equation*}\label{eq:additivity}
    \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
\end{equation*}
Then, once $C$ is fixed, the validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses; their validation losses are biased estimates of the loss function.

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as the training sample. The validation losses from all such \sq{folds} are averaged to produce the \emph{cross-validation loss}.

Cross-validation loss accurately reproduced the relationships between the excess losses of the estimators, with somewhat greater variability (\figref{1}{Row 6}).

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned to the stimulus onset for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low. The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051, mean 0.018 (\figref{2}{F} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by non-negative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients. The red line indicates the mean correlation coefficient of 0.020.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

In these dense populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimators that explicitly modeled the partial correlations between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Methods for details). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<10^{-2}$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

@@

We also examined the relationship of the correlation structure to the differences in orientation preference and to the physical distances between the cell pairs (Fig.\;\ref{fig:6}).  The partial correlations of the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation (\figref{6}{A}) than the sample correlations. They also decreased more rapidly with physical distance, both laterally  (\figref{6}{B}) and in depth (\figref{6}{B}).

>>>>

We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best modeled as the combination of a sparse graph of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

Improved estimation of the correlation matrix is beneficial in itself. For example, improved estimates can be used to optimize  decoding of the population activity \cite{Friedman:1989}. But reduced estimation error is not the only benefit of regularization.  The estimation improvement is greatest when the structure imposed by regularization limits the solution space while still allowing representation of the dominant dependencies in the data. Therefore, finding the most efficient among many regularized estimators leads to insights about the system itself: the structure of the most efficient estimator is a parsimonious representation of the regularities in the data.

The advantages due to regularization increase with the size of the recorded population. With the advent of  big neural data \cite{Alivisatos:2013}, the search for optimal regularization schemes will become increasingly relevant in any model of population activity. Since optimal regularization schemes are specific to systems under investigation, the inference of functional connectivity in large-scale neural data will entail the search for optimal regularization schemes. Such schemes may involve combinations of heuristic rules and numerical techniques specially designed for given types of neural circuits.

@@

We then performed a cross-validated evaluation to establish which of the four regularized estimators was most efficient for representing the population activity of dense groups of neurons in mouse primary visual cortex recorded with high-speed 3D random-access two-photon imaging of calcium signals. In our data, the sample correlation coefficients were largely positive and low.  We found that the best estimate of the correlation matrix was $C_{\sf sparse+latent}$.  This estimator revealed a sparse network of partial correlations (\sq{interactions}), between the observed neurons; it also inferred latent units exerting linear effects on the observed neurons. We analyzed these networks of partial correlations and found the following: Whereas significant noise correlations were predominantly positive, the inferred interactions had a large fraction of negative values possibly reflecting inhibitory circuitry.  Moreover, we found that these interactions exhibited a stronger relationship to the physical distances and to the differences in preferred orientations than noise correlations. In contrast, the inferred negative interactions were less selective.

@@

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an independent unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix},
\begin{equation}\label{eq:sample}
	C_{\sf sample} = \left[\mbox{cov}\left(x(t)\right)\right]_{t\in T} =\frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
\end{equation}
where $n$ is the size of $T$, yields an unbiased estimate of the covariance matrix so that $\E{C_{\sf sample}}=\Sigma$.

When the mean is estimated from the same sample, $C_{\sf sample}$ is biased toward zero.  However, in all the cases when the unbiasedness is required in our study, the mean will be estimated from a separate sample.

@@

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominently in probablistic \emph{graphical modelling} wherein the joint distribution is explained by sets of two-way interactions \cite{Whittaker:1990}. For the multivariate Gaussian distribution, zero partial correlations indicate conditional independence of the pair, implying a lack of direct interaction \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful as indicators of conditional independence (lack of functinonal connectivity) between pairs of neurons.

Pairwise partial correlations are closely related to the elements of the \emph{precision matrix}, \ie the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1990}. Zero elements in the precision matrix signify zero partial correlation between the two variables. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

As the size of recorded neuronal populations increases, without substantial increases in recording durations, the \emph{condition number} of the sample covariance matrix also increases, making the partial correlation estimates \emph{ill-conditioned}: small errors in covariance estimates translate into much larger errors in partial correlations. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization} \cite{Ledoit:2004,Schafer:2005}.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure01.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
    {\bf Row 1.} Graphical representations of the target estimates of the four respective regularized covariance matrix estimators.  Recorded neurons are represented by green spheres and latent units by light-shaded spheres.  Edges represent non-zero partial correlations, \ie \sq{interactions}.
        {\bf Row 1, A}.  For estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear dependencies. 
        {\bf  Row 1, B.} For estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), representing systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. For estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from a sparse set of  linear associations between the observed units. 
        {\bf  Row 1, D}.  For estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to sparse associations between the recorded cells \emph{and} with several latent units. \\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with four latent units, {\bf C.}  a correlation matrix with 67\%  off-diagonal zeros in its inverse, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-3 (\ie three latent units) matrix and a sparse matrix with 76\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=500$ drawn from simulated random processes with respective correlation matrices shown in Row 2.  The structure of the sample correlation matrix is difficult to discern.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:loss}) for the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with structure that matches the true model converged to zero faster than the other estimators. Some exceptions occur when the sample is too small to reveal the structure in the data. In these cases estimators with simpler structures can be more efficient than estimators with true structure.
\\
{\bf Row 6:} Cross-validation losses for the five estimators relative to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:1}
\end{FPfigure}

The target estimate of estimator $C_{\sf diag}$ is the diagonal matrix $D$ containing estimates of neurons' variances. Regularization is achieved by linear \emph{shrinkage} of the unbiased estimate $C_{\sf sample}$ toward $D$ controlled by the scalar \emph{shrinkage intensity} parameter $\lambda \in [0, 1]$:
\begin{equation}\label{eq:c-diag}
C_{\sf diag} = (1-\lambda) C_{\sf sample} + \lambda D
\end{equation}
The structure imposed by $C_{\sf diag}$ favors (performs better  on) populations   with no linear associations between the neurons (\figref{1}{Row 1, A}).  If a predominant majority of sample correlations are spurious, $C_{\sf diag}$ is expected to be more efficient than other estimators.

@@

Estimator $C_{\sf sparse+latent}$ is obtained by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component: 
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a low-rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. See Methods for a more detailed explanation. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (\figref{1}{Row 1, D}) \cite{Chandrasekaran:2010,Ma:2013}.

@@

\paragraph{Simulation}
We next demonstrate how the most efficient among different regularized estimators can reveal the structure of correlations.
We constructed four families of $50\times 50$ covariance matrices, each with structure that matched one of the four regularized estimators (\figref{1}{\,Row 2, A--D} and Methods).  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=500$), the sample correlation matrices were contaminated by sampling noise and their underlying structures were difficult to discern (\figref{1}{\,Row 3}).

@@

This loss function is related to the multivariate normal log-likelihood function $L(\Sigma|C_{\sf sample})$,
\begin{equation}
	\loss{X,Y} \equiv -\ln(2\pi) - \frac 2 {pn} L(X|Y)
\end{equation}
Normalization by $p$ and $n$ makes the values of the loss function comparable across different population sizes and sample sizes.

@@

The choice of the normal loss function is motivated, in part, by mathematical convenience. We expect that the main conclusions of our study will not change qualitatively with other well behaved loss functions, such as Stein's entropy loss or quadratic loss \cite{James:1961, Fan:2008, Ledoit:2004, Schafer:2005}.

We drew 30 independent samples of sizes $n=250$, 500, 1000, 2000, and 4000 from each model and computed the \emph{excess loss} $\loss{C,\Sigma}-\loss{\Sigma,\Sigma}$ for each of the five estimators.  The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  All the regularized estimators produced better estimates (lower excess loss) than the sample covariance matrix.  However, estimators whose structure matched the true model outperformed the other estimators (\figref{1}{\,row 5}).  Note that when the ground truth had zero correlations (Column A), $C_{\sf factor}$ performed equally well to $C_{\sf diag}$ because it correctly identified zero factors and only estimated the individual variances. Similarly, when the number of latent units was zero (Column C), $C_{\sf sparse+latent}$ performed equally well to $C_{\sf sparse}$ because it correctly identified zero latent units.

@@

With empirical data, the ground truth, $\Sigma$, is not accessible and excess loss cannot be computed directly. Instead, the loss function may be estimated from the sample through \emph{validation}.  In a validation procedure, an independent \emph{testing sample} is set aside to compute an additional sample covariance estimate $C_{\sf sample}^\prime$ to validate the estimate $C$ computed from the  \emph{training sample}.   The estimate $C_{\sf sample}^\prime$ is used to compute the \emph{validation loss} $\loss{C,C_{\sf sample}^\prime}$, which approximates the loss  $\loss{C,\Sigma}$.

The normal loss (Eq.~\ref{eq:loss}) is particularly suitable because it is additive in its second argument, \ie
\begin{equation*}\label{eq:additivity}
    \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
\end{equation*}
Then, once $C$ is fixed, the validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

The property of additivity does not hold for other popular loss functions such as Stein's entropy loss or various quadratic losses; their validation losses are biased estimates of the loss function. For these loss functions, cross-validation does not generally evaluate the discrepancy from truth.

The loss function above assumes a single covariance matrix across all conditions in the dataset.  However, neuroscientists often estimate a common correlation matrix across multiple stimulus conditions when the means and variances of responses are conditioned on the stimulus while correlations are assumed constant across all conditions \cite{Vogels:1989, Ponce:2013}. In this case, the stimulus-dependent means and variances are estimated from the training dataset along with a common correlation matrix. The adaptation of the normal loss function to this case is described in Methods.

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the testing sample with the other $K-1$ serving as the training sample. The validation losses from all such \sq{folds} are averaged to produce the \emph{cross-validation loss}.

Cross-validation loss accurately reproduced the relationships between the excess losses of the estimators (\figref{1}{Row 6}).

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low. The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean of 0.018 (\figref{2}{F} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by non-negative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients. The red line indicates the mean correlation coefficient of 0.018.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

In these densely sampled populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimators that explicitly modeled the partial correlations between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Methods). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<10^{-2}$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

@@

We also examined the relationship of the correlation structure to the differences in orientation preference and to the physical distances between the cell pairs (Fig.\;\ref{fig:6}).  The partial correlations from the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation than the sample correlations (\figref{6}{A}). They also decreased more rapidly with physical distance, both laterally (\figref{6}{B}) and in depth (\figref{6}{B}). The faster dropoff may be a trivial result of regularization: $L_1$-shrinkage of the inverse covariance reduces small partial correlations by a greater fraction than large partial correlations.
